Supplement to Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes

### 2020-12-29 manuscript/secs/introduction.tex
fix intro

Colorectal cancer (CRC) is a common malignancy in the United States.
Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
To improve overall survival of patients with distant disease, new therapeutic approaches are needed.

Immunotherapy in CRC is a promising new treatment option for patients with distant disease.
Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
protein 4 (CTLA-4) \citep{Le2015, Overman2018}.
However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \citep{Ganesh2019}.
Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
Epitopes derived from endogenous retroviral peptides expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \citep{Boller1997, Rooney2015}.
In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.

The cancer specificity of putative HERV peptides makes them promising biomarkers for and targets of immune therapies.
A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.

>>>>

Colorectal cancer (CRC) is a common malignancy in the United States.
Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
However, early detection relies on screening strategies in which nearly a third of adults do not participate as recommended \citep{CDC2019}.
Furthermore, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
To improve detection and treatment of patients with colorectal cancer, new diagnostic and therapeutic approaches are needed.

Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, but novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
Unfortunately, only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity are required to help extend the benefit of checkpoint blockade.
Endogenous retroviral elements expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can still be transcribed and translated, with some protein products even processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Rooney2015}.
In healthy tissues, DNA methylation and histone modification can silence transcription of HERV proviruses, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.

The HERV transcriptome is a promising source of biomarkers for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression.
A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
Proviruses are generally absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation and quantification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019, Grabski2020}.
Here, we extend advances in the characterization of HERV transcripts to discover novel CRC specific expression patterns in human colorectal transcriptomes.
We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
We tested this hypothesis in a large cohort of healthy, tumor adjacent, and tumor RNA sequencing data and found several HERV proviruses over-expressed in tumors but many that had greater expression in healthy tissue.
